### 2021-09-28 papers/thesis/resubmitted/abstract/abstract.tex
Resubmitted version of thesis to track the changes for the corrections. Fixed typo in SH clenshaws.

We begin with an introduction to sparse spectral methods via viewing spherical harmonics as multidimensional orthogonal polynomials in $x$, $y$, $z$. We explain how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how vector spherical harmonics in $x$, $y$ and $z$ can be used as an orthogonal basis for vector-valued functions, yielding similar banded-block-banded gradient and divergence operators.

@@

Piecing together the techniques used thus far, we present a new orthogonal polynomial basis and sparse spectral method for the spherical cap, complete with the same observation of the guaranteed sparsity of operators. The motivation is for one to use spherical caps bands as in a spectral element method for the sphere, with many applications in meteorology and astrophysics -- in particular, as a potential replacement of the spherical harmonics approach currently in use at ECMWF (which is predicted to become too costly due to a parallel scalability bottleneck arising from the global spectral transform).

>>>>

We begin with an introduction to sparse spectral methods via viewing spherical harmonics as multidimensional orthogonal polynomials in $x$, $y$, and $z$. We explain how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how vector spherical harmonics in $x$, $y$, and $z$ can be used as an orthogonal basis for vector-valued functions, yielding similar banded-block-banded gradient and divergence operators.

@@

Piecing together the techniques used thus far, we present a new orthogonal polynomial basis and sparse spectral method for the spherical cap, complete with the same observation of the guaranteed sparsity of operators. The motivation is for one to use spherical caps bands as in a spectral element method for the sphere, with many applications in meteorology and astrophysics -- in particular, as a potential replacement of the spherical harmonics approach currently in use at ECMWF, which is predicted to become too costly due to a parallel scalability bottleneck arising from the global spectral transform.

### 2021-09-28 papers/thesis/resubmitted/acknowledgements/acknowledgements.tex
Resubmitted version of thesis to track the changes for the corrections. Fixed typo in SH clenshaws.

I would like to thank the MPE CDT staff for the knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for his guidance, assistance and general enthusiasm (it meant a lot). And to all my fellow \enquote{Charlies} -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

>>>>

I would like to thank the MPE CDT staff for the knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for your guidance, assistance and general enthusiasm. And to all my fellow \enquote{Charlies} -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

### 2021-09-28 papers/thesis/resubmitted/ch2-spherical-harmonics/spherical-harmonics.tex
Resubmitted version of thesis to track the changes for the corrections. Fixed typo in SH clenshaws.

A multivariate analogue of Clenshaw's recursive algorithm for evaluation of a function expanded in an OP basis has been established for methods on other domains in 2D such as the triangle \cite{olver2019triangle}, and we proceed similarly here. We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The multidimensional \textit{Clenshaw algorithm} is as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+2} = \vec{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0.
}

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+2} = \vec{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices (note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).

>>>>

A multivariate analogue of Clenshaw's recursive algorithm for evaluation of a function expanded in an OP basis has been established for methods on other domains in 2D such as the triangle \cite{olver2019triangle}, and we proceed similarly here. We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The multidimensional \textit{Clenshaw algorithm} is as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0.
}

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } V = V(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices (note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).

### 2021-09-28 papers/thesis/resubmitted/ch4-spherical-caps/spherical-caps.tex
Resubmitted version of thesis to track the changes for the corrections. Fixed typo in SH clenshaws.

Just as for the spherical harmonics framework for the whole sphere in \bsrefchapter{CHAPTER:sphericalharmonics}, we consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator). Recall that we write the spherical coordinates as
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $ \rho(z) := \sqrt{1-z^2}$, and that the Laplace--Beltrami operator is then
\bseqn{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for $\xvec := (x,y,z) \in \R^3$. We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see \bsreftheorem{theorem:sc:sparsityofdifferentialoperators}. Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, a la the ultraspherical spectral method we investigated in \bsrefchapter{CHAPTER:diskslice} for disk-slices and trapeziums (see also \cite{snowball2019sparse}), as well as for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle}, and the related work on sparse discretisations involving Jacobi polynomials on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}. Just as we proceeded in \bsrefchapter{CHAPTER:diskslice} for the disk-slice case in 2D (see also \cite{snowball2019sparse}), we will use an integration-by-parts argument to deduce the sparsity structure.

@@

The spherical cap can be thought of as a higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). Recently, a procedure for defining OPs in two variables on quadratic curves, such as ellipses and parabolas, has been presented \cite{olver2021orthogonal}, where the unit circle is one of the five classes of curve that can be generalised to give any curve in the plane $\R^2$. One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$:
\begin{definition}[\cite{olver2021orthogonal}]\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta) \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta)
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

@@

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, on the triangle \cite{olver2018recurrence}, and on the disk-slice \cite{snowball2019sparse}. The operators we define here are for partial derivatives with respect to the spherical coordinates $(\varphi, \theta)$, so that we can more easily apply the operators to PDEs on the surface of a sphere (for example, surface Laplacian operator in the Poisson equation). With the OP ordering by Fourier mode $k$ defined in \bsrefeqns{eqn:sc:OPdefNka}{eqn:sc:OPdefNa} these rotationally invariant operators are block-diagonal, meaning simple and parallelisable practical application.

@@

Denote the condition number of the matrix $A$: 
\bseqn{
	\kappa(A) := \norm{A}_2 \norm{A^{-1}}_2,
}
which encodes how accurate the solution to $A \vc x = \vc b$ is under perturbation. Since the Laplacian is block-diagonal one can determine the condition number from the condition numbers of each block on the diagonal,  and hence deduce the accuracy in solving the Poisson equation, where graphical representations are seen in \bsreffig{fig:sc:conditionnumbers} for the $\Delta^{(1)}_W$ operator matrix. As the degree $N$ increases, the condition numbers of these blocks becomes grows  algebraically fast (at least, the condition number for the first block where $m=0$ does, which will be the largest for each $N$). Fortunately however, by applying a trivial diagonal preconditioning matrix similar to that seen for the ultraspherical method \cite{olver2013fast}, we can significantly bound this growth of the condition numbers -- this is also seen in \bsreffig{fig:sc:conditionnumbers}. The preconditioning matrix chosen is $P^{-1}$, where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$.

@@

Define $X := (J_x^{(0)})^\top$, $Y := (J_y^{(0)})^\top$, $Z := (J_z^{(0)})^\top$. We can obtain the matrix operator for the variable-coefficient function $v(x,y,z)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices $X, Y, Z$, yielding an operator matrix of the same dimension as the input Jacobi matrices a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bigscopNo$ space by $v(X, Y, Z)$. In matrix-vector notation, our system hence becomes:
\bseqn{
    (\Delta_W^{(1)} + k^2 \:T^{(0)\to(1)} \: V \: T_W^{(1)\to(0)}) \: \mathbf{u} = \mathbf{f}
}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in \bsreffig{fig:sc:sparsity} with $v(x,y,z) = zxy^2$ as an example. In \bsreffig{fig:sc:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in \bsrefeqn{eqn:sc:helmholtz} in the spherical cap $\Omega$, with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. In \bsreffig{fig:sc:helmholtz} we also show the norms of each block of calculated coefficients for the approximation of the solution to the inhomogeneous variable-coefficient Helmholtz equation with various $k$ values. Here, we use $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve spectral convergence.

>>>>

Just as for the spherical harmonics framework for the whole sphere in \bsrefchapter{CHAPTER:sphericalharmonics}, we consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator). Recall that we write the spherical coordinates as
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $ \rho(z) := \sqrt{1-z^2}$, and that the Laplace--Beltrami operator is then
\bseqn{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for $\xvec := (x,y,z) \in \R^3$. We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see \bsreftheorem{theorem:sc:sparsityofdifferentialoperators}. Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, \`a la the ultraspherical spectral method we investigated in \bsrefchapter{CHAPTER:diskslice} for disk-slices and trapeziums (see also \cite{snowball2019sparse}), as well as for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle}, and the related work on sparse discretisations involving Jacobi polynomials on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}. Just as we proceeded in \bsrefchapter{CHAPTER:diskslice} for the disk-slice case in 2D (see also \cite{snowball2019sparse}), we will use an integration-by-parts argument to deduce the sparsity structure.

@@

The spherical cap can be thought of as a higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). Recently, a procedure for defining OPs in two variables on quadratic curves, such as ellipses and parabolas, has been presented \cite{olver2021orthogonal}, where the unit circle is one of the five classes of curve that can be generalised to give any curve in the plane $\R^2$. One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$.
\begin{definition}[\cite{olver2021orthogonal}]\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta) \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta)
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

@@

The incrementing and decrementing of parameters as seen here is analogous to other to that seen in \bsrefchapter{CHAPTER:diskslice} (see also \cite{snowball2019sparse}), and we further reiterate that it is also seen for other well known orthogonal polynomial families' derivatives -- for example the Jacobi polynomials on the interval (as seen in the NIST Digital Library of Mathematical Functions \cite[(18.9.3)]{DLMF}) and on the triangle \cite{olver2018recurrence}. The operators we define here are for partial derivatives with respect to the spherical coordinates $(\varphi, \theta)$, so that we can more easily apply the operators to PDEs on the surface of a sphere (for example, surface Laplacian operator in the Poisson equation). With the OP ordering by Fourier mode $k$ defined in \bsrefeqns{eqn:sc:OPdefNka}{eqn:sc:OPdefNa} these rotationally invariant operators are block-diagonal, meaning simple and parallelisable practical application.

@@

Denote the condition number of the matrix $A$: 
\bseqn{
	\kappa(A) := \norm{A}_2 \norm{A^{-1}}_2,
}
which encodes how accurate the solution to $A \vc x = \vc b$ is under perturbation. Since the Laplacian is block-diagonal one can determine the condition number from the condition numbers of each block on the diagonal,  and hence deduce the accuracy in solving the Poisson equation, where graphical representations are seen in \bsreffig{fig:sc:conditionnumbers} for the $\Delta^{(1)}_W$ operator matrix. As the degree $N$ increases, the condition numbers of these blocks grows algebraically fast (at least, the condition number for the first block where $m=0$ does, which will be the largest for each $N$). Fortunately however, by applying a trivial diagonal preconditioning matrix similar to that seen for the ultraspherical method \cite{olver2013fast}, we can significantly bound this growth of the condition numbers -- this is also seen in \bsreffig{fig:sc:conditionnumbers}. The preconditioning matrix chosen is $P^{-1}$, where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$.

@@

Define $X := (J_x^{(0)})^\top$, $Y := (J_y^{(0)})^\top$, $Z := (J_z^{(0)})^\top$. We can obtain the matrix operator for the variable-coefficient function $v(x,y,z)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices $X, Y, Z$, yielding an operator matrix of the same dimension as the input Jacobi matrices \`a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bigscopNo$ space by $v(X, Y, Z)$. In matrix-vector notation, our system hence becomes:
\bseqn{
    (\Delta_W^{(1)} + k^2 \:T^{(0)\to(1)} \: V \: T_W^{(1)\to(0)}) \: \mathbf{u} = \mathbf{f}
}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in \bsreffig{fig:sc:sparsity} with $v(x,y,z) = zxy^2$ as an example. In \bsreffig{fig:sc:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in \bsrefeqn{eqn:sc:helmholtz} in the spherical cap $\Omega$, with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. In \bsreffig{fig:sc:helmholtz} we also show the norms of each block of calculated coefficients for the approximation of the solution to the inhomogeneous variable-coefficient Helmholtz equation with various $k$ values. Here, we use $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve spectral convergence.
